---
title: "Plotting with miamiplot"
author: "Julie White"
date: "` r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{miamiplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(miamiplot)
library(dplyr)
library(RColorBrewer)
library(viridisLite)
library(ggplot2)
```

This document acts as a tutorial for using the miamiplot package to create Miami
plots for GWAS or EWAS results. Use `?FunctionName` in the R console to get the 
complete documentation of a given function.

# Data
The `miamiplot` package includes a data.frame called `gwas_results` with 
simulated results from two studies for 30,000 SNPs on 22 chromosomes. 
```{r}
str(gwas_results)
```

# Plotting
## Assigning values to to the upper and lower sections of the plot
Say you wanted to plot the GWAS from study A and wanted to separate
the plot by beta values, with the variants with positive beta values in the 
upper plot. While we're at it, we can specify y-axis label names.
```{r, error = TRUE, purl = FALSE}
ggmiami(data = gwas_results[which(gwas_results$study == "A"),], 
        split_by = "beta", split_at = 0, p = "pval", 
        upper_ylab = "Positive beta values",
        lower_ylab = "Negative beta values")
```
It seems like most of the lower-value SNPs in the peak on chromosome 15 have 
positive beta values. 

Say instead you wanted to plot the results from study "A" in the upper plot and
study "B" in the lower plot.

```{r}
ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A", lower_ylab = "Study B")
```

## Changing the significance lines
The suggestive and genome-wide significance lines can either be turned off or 
given different p-values. In this example we'll turn off the suggestive line and
plot the genome-wide line at a very small value: 5 x 10^-15
```{r}
ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A",lower_ylab = "Study B", suggestive_line = NULL, 
        genome_line = 5e-15)
```

## Changing colors
The default chromosome colors are alternating black and grey. If you'd prefer to
supply your own alternating colors, you can do so using color names, e.g. 
`chr_colors = c("blue", "orange")`, hexidecimal names, e.g. 
`chr_colors = c("#d8b365", "#5ab4ac")`, or a list of colors equal to the number 
of chromosomes in your plot. 

Here, using \code{brewer.pal}
```{r}
mycolors <- colorRampPalette(brewer.pal(12, "Paired"))(22)

ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A", lower_ylab = "Study B", chr_colors = mycolors, 
        genome_line_color = "black", suggestive_line_color = "#A9A9A9")

rm(mycolors)
```

## Adding labels
To simply label the top 5 hits on each plot, all you need to do is supply the 
name of the column from which we should draw the lables. Here, we'll label the 
top 5 from each plot section with the rsid.
```{r}
ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A", lower_ylab = "Study B", 
        hits_label_col = "rsid")
```

You can also label based on two columns, and change the number of labels: 
```{r}
ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A", lower_ylab = "Study B", 
        hits_label_col = c("rsid", "beta"), top_n_hits = 10)
```
This can get a bit messy. 

If you would instead like to supply a list of which items to label, you can 
use \code{hits_label} to supply the list, and \code{hits_label_col} to
specify where these values will be found.
```{r}
ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A", lower_ylab = "Study B", 
        hits_label = c("rs19142", "rs27017", "rs19103", "rs26991", "rs29240"),
        hits_label_col = "rsid")
```

### Custom labels
Say you wanted to be more specific and provide your own dataframe containing the
labelling information. This requires a little front-end work on your part, but 
gives you more control in case the above two options aren't producing what you
want. 

First, it's helpful to have access to the actual data being plotted, so that you
don't have to go about calculating relative genomic position on your own. We can
do this using the `prep_miami_data` function from this package.
```{r}
plot_data <- prep_miami_data(data = gwas_results, split_by = "study", 
                             split_at = "A", p = "pval")
```

Next, say we wanted to identify top peak from each chromosome and plot the five
with the lowest p-values, while labelling them with rsid and beta value.
```{r}
# Study A
studyA_labels <- plot_data$upper %>%
  group_by(chr) %>%
  arrange(desc(logged_p)) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(label = paste0(rsid, "\n", beta)) %>%
  select(rel_pos, logged_p, label) %>%
  arrange(desc(logged_p)) %>%
  slice(1:5)

# Study B
studyB_labels <- plot_data$lower %>%
  group_by(chr) %>%
  arrange(desc(logged_p)) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(label = paste0(rsid, "\n", beta)) %>%
  select(rel_pos, logged_p, label) %>%
  arrange(desc(logged_p)) %>%
  slice(1:5)
```

Now for the plot.
```{r}
ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A", lower_ylab = "Study B", 
        upper_labels_df = studyA_labels, lower_labels_df = studyB_labels)
```
You can also specify labels for only one plot. But you cannot mix label methods
```{r, error = TRUE, purl = FALSE}
ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A", lower_ylab = "Study B", 
        upper_labels_df = studyA_labels, hits_label_col = c("rsid", "beta"))
```

## Highlighting SNPs
If you wanted to highlight specific points, you can do so. 

For the upper highlighted SNPs, I'm going to highlight everything withih 100 bp
of the two genome-wide significant peaks.
```{r}
# Get the position of the two peaks +- 100 bp.
studyA_highlight_pos <- plot_data$upper %>%
  filter(pval < 5e-8) %>%
  group_by(chr) %>%
  filter(pval == min(pval)) %>%
  summarise(start = rel_pos - 100, end = rel_pos + 100) %>%
  select(-chr) %>%
  apply(., 1, function(x) x["start"]:x["end"]) %>%
  as.vector()

# Find which rsids match these SNPs
studyA_highlight_rsid <- plot_data$upper %>%
  mutate(in_peak = case_when(rel_pos %in% studyA_highlight_pos ~ "Yes", 
                             TRUE ~ "No")) %>%
  filter(in_peak == "Yes") %>%
  select(rsid)
```

Add to plot
```{r}
ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A", lower_ylab = "Study B", 
        upper_labels_df = studyA_labels, lower_labels_df = studyB_labels, 
        upper_highlight = studyA_highlight_rsid$rsid, 
        upper_highlight_col = "rsid")
```

You can also change the highlighting color with either names or hex codes 
using `upper_highlight_color` and `lower_highlight_color`. You can even specify specific colors for each point, if you wanted. I'll give an
example of making the color based on the p-value of the SNP using viridis 
colors.
```{r}
studyA_highlight_rsid <- plot_data$upper %>%
  mutate(in_peak = case_when(rel_pos %in% studyA_highlight_pos ~ "Yes", 
                             TRUE ~ "No")) %>%
  filter(in_peak == "Yes") %>%
  arrange(logged_p) %>%
  mutate(color = plasma(n = nrow(.), direction = -1)) %>%
  select(rsid, logged_p, color)
```

Add to plot
```{r}
ggmiami(data = gwas_results, split_by = "study", split_at = "A", p = "pval", 
        upper_ylab = "Study A", lower_ylab = "Study B", 
        upper_labels_df = studyA_labels, lower_labels_df = studyB_labels, 
        upper_highlight = studyA_highlight_rsid$rsid, 
        upper_highlight_col = "rsid", 
        upper_highlight_color = studyA_highlight_rsid$color)
```

## Saving the plot 
Since the plot produced is a patchwork of ggplot2 objects, it can be saved using ggsave.
```{r}
p <- ggmiami(data = gwas_results, split_by = "study", split_at = "A", 
             p = "pval", upper_ylab = "Study A", lower_ylab = "Study B", 
             upper_labels_df = studyA_labels, lower_labels_df = studyB_labels, 
             upper_highlight = studyA_highlight_rsid$rsid, 
             upper_highlight_col = "rsid", upper_highlight_color = "magenta")

ggsave(filename = "ExampleMiamiPlot.png", device = "png", width = 8, 
       height = 6, units = "in")
```

# A few notes
* The default values for chromosome, position and p-value are `chr`, `pos`, and
`p`, respectively. If you're creating a miami plot and your column names are
different, you'll have to pass the column names to the `chr = `, `pos = `, and 
`p = ` arguments. See help(ggmiami) for details. 
* The chromosome column must be numeric. If you have "X," "Y," or "MT" 
chromosomes, you'll need to rename these 23, 24, 25, etc. You can modify the
source code (e.g. `fix(ggmiami)`) to change the line designating the axis tick
labels in the upper plot `labels = plot_data$axis$chr` to set this to whatever
you'd like it to be. 

